\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-08-29 coqimpl.tex
Save.

However, "FRecursor" provides barely enough functionality for manipulating (extensible) inductive type, and it is far from satisfactory to be a legitimate programming interface---for example, the necessary tactic "inversion", "discriminate", "injection" is not applicable on (extensible) inductive type.

To resolve this problem, we notice that, the power of "FRecursor" comes from its distinction from the mundane fields and its ability to stab through the abstraction of "FInductive". However, the latter is quite similar on how "Overridable .. pins {..}" can see through the abstraction of former pinned overrdiable fields. This inspires us to generalize pins to not only apply to overrdiable fields, but also "FInductive" and "FRecursor" fields.

\textbf{Overridable pins Inductive Type and FRecursor}.

Also, when pins "FInductive" and "FRecursor", we can have the primitive coq-providing recursor and judgemental equality for computation ability for "FRecursor".

Upon this design, it is natural to make "Overridable" fields to provide ``automatic overriding'' as it is likely the inherited tactic is able to provide a proof even in the context with the extended inductive type.

But a question arises: do we consider an ``automatic overridden'' field \textit{inheriting} the parent field or \textit{overriding} the parent field? 
% TODO: Mention still needing total recursor, for exhaustiveness checking

@@

% Following needs a rewrite

\textbf{Propositional Partial Recursor.} This section is introducing the concept of \textit{partial recursor}. The motivations are various: (a) we want to have some semantic characterization of extensible inductive type. (b) We want to show that, some tactics reasoning like "inversion" is still valid even in extensible inductive type. (c) More generally speaking, notice that, for both "FRecursor" and "Closing Fact", they are not necessarily ``\textit{conservative extension}''---they will possibly disrupt arbitrary extension of inductive type. For example, for \cref{fig:STLC-example}, once we proved \mintinline{Coq}{var_not_value : ∀ i, ~ value (tm_var i)} in the "Family STLC" either using "FRecursor" or "Closing", then any children family of "STLC" cannot extend a new constructor \mintinline{Coq}{vvar: ∀ i, value (tm_var i)}. We want to see what leads to a conservative extension---and the partial recursor is helpful for this question.

% This raise an interesting question: what propositions are conservative extension? 
For example, apparently we know injectiveness of constructor, (i.e. \mintinline{Coq}{STLC.tt ≠ STLC.ff}) must be conservative because it holds for all extension of the inductive type "STLC.tm". But is there a \textbf{``best'' proposition} (i.e. the proposition that everything the proposition derives are exactly conservative extension)?

Since injectiveness of the constructors are derived by recursor in vanilla inductive type, an educated guess would be related to recursor. But we know the original recursor won't work because the mere existence of the original recursor will break extensibility. Thus we need to modify it.
The simplest idea is to allow partiality---i.e., returning
\mintinline{Coq}{option T} instead of the original \mintinline{Coq}{T}.

We call this \textit{propositional partial recursor}.\YZ{What is propositional about partial recursors? I suppose it has to do with the computational axioms?}\EDJreply{Yes. I want to emphasize the equality used by computational axiom is not judgemental equality. So Coq cannot automatically compute/reduce when a partial recursor applied to a term. This reduction can only be semi-automatically done by using [rewrite tactic and the computational axiom].}
For example, the propositional partial recursor for \mintinline{Coq}{B.b} is\YZ{Mention that partial recursors can be automatically generated}\EDJreply{Added below please check.}

\begin{minted}[escapeinside=@@]{Coq}
(* Inside Family B *)
FInductive b : Set := tt : b | ff : b.
b_prec : forall R, option R -> option R -> b -> option R.@\YZ{Not clear what it takes to define/prove partial recursors. The reader needs to be told if they require breaking abstraction like recursors do.}\EDJreply{What do you mean? recursor doesn't break abstraction but break extensibility. Below mentioned partial recursor doesn't break extensibility}@
\end{minted}
with appropriate computational axioms. This partial recursor differs from original recursor only by decorating an "option" with the return type thus possible to be automatically generated (and the case is similar for those computational axiom). The partiality doesn't break
extensibility (i.e. all future extensions of "b" can support this
"b_prec") because "b_prec" can just return "None" when bump into extended constructor in the future.

The first thing to check is that, if partial recursor is theoretically enough for injectiveness of the constructor. The rough idea to achieve this
is to use "b_prec" to reflect our special inductive type into vanilla
inductive type, and use the original "injection" and "discriminate"
tactics. For example, we want to prove \mintinline{Coq}{B.tt ≠
B.ff}, we simply 
\begin{minted}{Coq}
Definition reflect : B.b → option bool 
                   := b_prec (fun _ => bool) (Some true) (Some false).
Definition tt_neq_ff : B.tt = B.ff → False.
(* Because B.tt = B.ff → reflect B.tt = reflect B.ff 
      → Some true = Some false    (by computational axiom)
      → False           (by discriminate of bool and injection of Some) *)
\end{minted}

Even though we cannot prove propositional partial recursor is \textit{the best},
the above "reflect" function implies that propositional partial
recursor is a good \textit{extensional characterization} of (non-indexed) extensible inductive type (with certain constraints)---because we can ``embed'' vanilla inductive types into
types supporting a partial recursor: "reflect" is actually a left
inverse of an injection "bool → self_B.b". A bit more formally, and
restricting our focus to non-indexed inductive types:

\subsection{Limitation}
We list some of the current limitation of our implementation. We believe some of these limitations can be remedied by more engineering efforts.
\begin{itemize}
  \item We don't support the full power of inductive facility in Coq, that includes mutual inductive type, notation for inductive type and parameters in the inductive type (we only support indexed inductive type right now)
  \item We don't support mutual fixpoint definition
  \item We don't support incorporating "Hint" Vernacular commands, and thus the users cannot specify the database hints
  \item Unicode (i.e. "Require Import Coq.Unicode.Utf8.") is not compatible with our plugin 
  \item We don't consider the compatibility of universe polymorphism mechanism with our plugin
\end{itemize}

>>>>

However, "FRecursor" provides barely enough functionality for manipulating our(extensible) inductive type, and it is far from satisfactory to be a legitimate programming interface. (1) We do not yet have any computational information using this "FRecursor" until its compilation into module, i.e. we cannot do computation with it during the construction of the family using the current compilation strategy. (2) The necessary tactic like "discriminate" and "injection" is not applicable on (extensible) inductive type.

\textit{Generating Scheme for Propositional Computation Axiom for FRecursor}. We support "FScheme" that our plugin will automatically generate(postulate) the computational ``axiom'' for the specified  into the family. During compilation, these ``axiom'' can be easily proved/instantiated with concrete/vanilla inductive type.

However, this method, though remedies problem (1) to some extent, is only introducing computational information \textbf{propositionally} but not judgementally. Thus, for example, during theorem proving, we might need the help of rewriting tactic on these computational ``axiom'' to justify the computation. And in the full dependent type theory, propositional equality is sometimes not enough and bringing a lot of coding challenges\footnote{For example, for "u : F a"， "v : F b", "u = v" is not even well-typed when "a = b" only propositionally hold. Transport is required during reasoning.}.

% To resolve (2), we can simply  Our following three sections will try to resolve these problems. 
% we notice that, the power of "FRecursor" comes from its distinction from the mundane fields and its ability to stab through the abstraction of "FInductive". However, the latter is quite similar on how "Overridable .. pins {..}" can see through the abstraction of former pinned overrdiable fields. This inspires us to generalize pins to not only apply to overrdiable fields, but also "FInductive" and "FRecursor" fields.

\textit{Overridable pins Inductive Type and FRecursor}. We notice that, the power of "FRecursor" comes from its distinction from the mundane fields and its ability to stab through the abstraction of "FInductive". However, the latter is quite similar on how "Overridable .. pins {..}" can see through the abstraction of former pinned overrdiable fields. This inspires us to generalize pins---we support pinning inductive type and "FRecursor". Once "FInductive" and "FRecursor" fields are pinned, we can stab through its abstraction and have the primitive coq-providing recursor and \textbf{judgemental equality} for the computational information for "FRecursor".

The compilation mechanism is simply make the underlying, compiled definition of inductive type and recursor included inside the parameters of the module, just like how the the concrete definition of overridable fields are included inside the parameters when pinned. This way, the computational information from vanilla Coq can be transimitted into the context.

% However, only overridable fields can pin the (extensible) inductive type and "FRecursor".

% Upon this design, it is natural to make "Overridable" fields to provide ``automatic overriding'' as it is likely the inherited tactic is able to provide a proof even in the context with the extended inductive type.

% But a question arises: do we consider an ``automatic overridden'' field \textit{inheriting} the parent field or \textit{overriding} the parent field? 
% TODO: Mention still needing total recursor, for exhaustiveness checking

% Following needs a rewrite
To achieve "injection" in problem (2), we can easily generate/postulate $n$ propositions for $n$ constructors, just like how we did for computational information. But for "discriminate" tactic, this way will generate $O(n^2)$ propositions.

What's more, to manipulate inductive type, it is not necessarily enough to only support "injection" and "discriminate" tactic (and their corresponding axioms). Recall in vanilla inductive type, we believe that eliminator is the strong-enough ``toolkit'' because the eliminator (with its computational information) is the \textit{universal property} of the inductive type---it extensionally describes the inductive type up to (certain) isomorphism. This universal property is much more powerful than "injection" and "discriminate" and maybe the most powerful tool we can have to reason about inductive type.

For example, given a new inductive type, to prove "discriminate" between two constructors, i.e., "c$_i$ .. $\neq$ c$_j$ .." for two different constructors, we simply use the eliminator to map each constructor to different natural number, assuming we can "discriminate" different natural number.\footnote{Rigorously, to avoid any suspicious circular reasoning, we can use large elimination to map them to $\top$ and $\bot$, and because $\exists a : ?, \top$ holds for the former not the later, thus $\top \neq \bot$.} "injection" can be supported in a similar way---we can simply let eliminator map each inductive type to the arguments of the constructors.

Thus it is natural for us to think if (an altered form of) eliminator can be postulated into the context, since apparently we cannot directly postulate the original eliminator here---because the sole existence of the eliminator prevents any possible future extension of the inductive type.

Thus we alter eliminators into \text{partial recursor}, such that (1) it will not hinder the future extension, and (2) it is still powerful enough to derive injectiveness like tactic "inject" and "discriminate". What's more, (3) we can actually prove this \text{partial recursor} is actually good enough because it can act as an \textit{extensional characterization} for extensible inductive type. We will call it \textit{propositional} because we only prpositionally postulate its computational informations in Coq.s

\textbf{Propositional Partial Recursor.} Partial recursor is simply adding partially into the signature of the eliminator, by wrapping each motive with "option". 
% This section is introducing the concept of \textit{partial recursor}. The motivations are various: (a) we want to have some semantic characterization of extensible inductive type. (b) We want to show that, some tactics reasoning like "inversion" is still valid even in extensible inductive type. (c) More generally speaking, notice that, for both "FRecursor" and "Closing Fact", they are not necessarily ``\textit{conservative extension}''---they will possibly disrupt arbitrary extension of inductive type. For example, for \cref{fig:STLC-example}, once we proved \mintinline{Coq}{var_not_value : ∀ i, ~ value (tm_var i)} in the "Family STLC" either using "FRecursor" or "Closing", then any children family of "STLC" cannot extend a new constructor \mintinline{Coq}{vvar: ∀ i, value (tm_var i)}. We want to see what leads to a conservative extension---and the partial recursor is helpful for this question.

% This raise an interesting question: what propositions are conservative extension? 
% For example, apparently we know injectiveness of constructor, (i.e. \mintinline{Coq}{STLC.tt ≠ STLC.ff}) must be conservative because it holds for all extension of the inductive type "STLC.tm". But is there a \textbf{``best'' proposition} (i.e. the proposition that everything the proposition derives are exactly conservative extension)?

% Since injectiveness of the constructors are derived by recursor in vanilla inductive type, an educated guess would be related to recursor. But we know the original recursor won't work because the mere existence of the original recursor will break extensibility. Thus we need to modify it.
% The simplest idea is to allow partiality---i.e., returning
% \mintinline{Coq}{option T} instead of the original \mintinline{Coq}{T}.

% We call this \textit{propositional partial recursor}.\YZ{What is propositional about partial recursors? I suppose it has to do with the computational axioms?}\EDJreply{Yes. I want to emphasize the equality used by computational axiom is not judgemental equality. So Coq cannot automatically compute/reduce when a partial recursor applied to a term. This reduction can only be semi-automatically done by using [rewrite tactic and the computational axiom].}
For example, the propositional partial recursor for \mintinline{Coq}{B.b} is\YZ{Mention that partial recursors can be automatically generated}\EDJreply{Added below please check.}

\begin{minted}[escapeinside=@@]{Coq}
(* Inside Family B *)
FInductive b : Set := tt : b | ff : b.
b_prec : forall R, option R -> option R -> b -> option R.@\YZ{Not clear what it takes to define/prove partial recursors. The reader needs to be told if they require breaking abstraction like recursors do.}\EDJreply{What do you mean? recursor doesn't break abstraction but break extensibility. Below mentioned partial recursor doesn't break extensibility}@
\end{minted}
with appropriate computational axioms. This partial recursor differs from original recursor only by decorating an "option" with the return type thus possible to be automatically generated (and the case is similar for those computational axiom). Thanks to the partiality, partial recursor doesn't break
extensibility (i.e. all future extensions of "b" can support this
"b_prec") because for example, "b_prec" can just return "None" when bump into extended constructor in the future.

The second thing to check is that, if partial recursor is theoretically enough for injectiveness of the constructor. The rough idea to achieve this
is to use "b_prec" to reflect our special inductive type into vanilla
inductive type, and use the original "injection" and "discriminate"
tactics. For example, we want to prove \mintinline{Coq}{B.tt ≠
B.ff}, we simply 
\begin{minted}{Coq}
Definition reflect : B.b → option bool 
                   := b_prec (fun _ => bool) (Some true) (Some false).
Definition tt_neq_ff : B.tt = B.ff → False.
(* Because B.tt = B.ff → reflect B.tt = reflect B.ff 
      → Some true = Some false    (by computational axiom)
      → False           (by discriminate of bool and injection of Some) *)
\end{minted}

Thirdly, even though we cannot prove partial recursor is \textit{the best} reasoning tool for extensible inductive type,
the above "reflect" function implies that propositional partial
recursor is a good \textit{extensional characterization} of (non-indexed) extensible inductive type (under certain constraints)---because we can ``embed'' vanilla inductive types into
types supporting a partial recursor: "reflect" is actually a left
inverse of an injection "bool → self_B.b". A bit more formally, and
restricting our focus to non-indexed inductive types:

Though partial recursor is provable by simply using "FRecursor", for convenience, we support using "FScheme" to automatic generate it. At the meantime, we implement a tactic "prec_discriminate" that when fed with the appropriate partial recursor, it can discriminate an absurd equation between two different constructors of extensible inductive type.

Finally, we should point out that, the mere postulation of partial recursor is \textit{conservative}. In other word, generally speaking, using "FRecursor" to prove theorems is not necessarily a ``\textit{conservative extension}''---the proved theorems might hinder future extensions of the inductive type. For example, for \cref{fig:STLC-example}, once we proved \mintinline{Coq}{var_not_value : ∀ i, ~ value (tm_var i)}\footnote{Achieved by three steps: first proving partial recursor using "FRecursor", then use partial recursor to prove "tm_var i $\neq$ tm_lam v body" and finally using "FRecursor" prove "var_not_value" by induction on "value"} in the "Family STLC" either using "FRecursor", then any children family of "STLC" cannot extend a new constructor \mintinline{Coq}{vvar: ∀ i, value (tm_var i)}. But partial recursor will not cause this problem.

@@

\subsection{Engineering Limitation}
We list some of the current limitation of our implementation. We believe some of these limitations can be remedied by more engineering efforts.
\begin{itemize}
  \item We don't support the full power of inductive facility in Coq, that includes mutual inductive type, notation for inductive type and parameters in the inductive type (we only support indexed inductive type right now)
  \item We don't support mutual fixpoint definition
  \item We don't support incorporating "Hint" Vernacular commands, and thus the users cannot specify the database hints
  \item Unicode (i.e. "Require Import Coq.Unicode.Utf8.") is not compatible with our plugin 
  \item We don't consider the compatibility of universe polymorphism mechanism with our plugin
\end{itemize}
